\beforeabstract
\prefacesection{Abstract}

This project examines the advantages and development of Machine Learning (ML) models designed to \textbf{explain the factors influencing property values in natural human language}.

The project is divided into two primary components: implementing eXtreme Gradient Boosting (XGBoost) to predict property values based on specific features (e.g. square metres, number of bedrooms) and utilising SHapley Additive exPlanations (SHAP) to articulate, in human language, the rationale behind the model's evaluation. 

The literature review underscores several benefits of eXplainable Artificial Intelligence (XAI) within the property market sector, highlighting the significance for end users receiving detailed, human-readable explanations of the factors affecting the model's judgement, as opposed to a mere score system, where the model's reasoning remains opaque. 

The chosen model for the task, XGBoost, is a gradient-boosting decision-making machine learning model that parallels the human evaluation of property values in the market by considering various attributes and generating an overall estimate based on the data provided during model training. Following the predictions of XGBoost, SHAP identifies the features that have increased or decreased the prediction values and quantifies their impact, ultimately generating explanatory text based on the influence of different features of the dataset.

\prefacesection{Acknowledgments}
I express my sincere gratitude to my project supervisor, Ross McLean, for his guidance throughout this project and for his assistance in resolving the various challenges I encountered along the way. His professionalism and dedication have been truly exceptional and I consider myself fortunate to have worked under his supervision.

I extend my thanks to the entire School of Computing, Engineering and Technology at the Robert Gordon University for the knowledge, support, and formative experiences that have shaped me into a better version of myself. I expressly acknowledge professor John N. Brown, whose teaching broadened my perspective and encouraged me to think critically and beyond conventional boundaries. 

I would like to dedicate this work to my mother and father, whose sacrifices and unwavering support have made it possible for me to reach this point; and to my brother, whose presence has been a constant source of strength. I am also very thankful to my aunt, who cared for me as if I were her own child.

Finally, my heartfelt thanks go to my girlfriend, Ivana, who stood with me throughout this glorious journey. Not for a single moment did she stop believing in me. Even in the moments when I was at my lowest, mentally and physically, and ready to give up, she refused to let me do so. Her encouragement and unwavering faith carried me to the finish line, and for that I will always be grateful.

\afterpreface \afterabstract